A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-05-09 papers/water-SAFT/paper.tex
Removed water-vapor density plot and moved written section to empirical parameters. Fixed directory paths in papers/water-SAFT/figs/ Tested everything on quipu.

% needsworknow is intended to prioritize stuff that needs fixing.
\newcommand{\needsworknow}[1]{\textcolor{red}{[\emph{#1}]}}

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, gloor2002saft, Jaqaman2004,
  clark2006developing, lischner2010classical}\needsworknow{ADD REST OF
  LIST HERE!}, capturing some of the qualitative behavior of water.
However, many of these are less sophisticated than those describing
the hard-sphere fluid, and a quantitatively accurate functional
describing water---which reproduces the temperature dependence of the
macroscopic surface tension, its decrease as curvature increases, and
its quantitative value---has not yet been published.  For instance,
the functional of Lischner \emph{et al}\cite{lischner2010classical}
treats the surface tension correctly, but can only be used at room
temperature.  A numer of classical density functionals have recently
been produced that are based on Statistical Associating Fluid Theory
(SAFT)\cite{gloor2002saft, clark2006developing}\needsworknow{add
  more}.  These functionals are based on a perturbative thermodynamic
expansion, and give a reasonable prediction for the temperature
dependence of the surface tension.  However, they also overestimate
the surface tension by \needsworknow{how much? significant factor, I
  think}, and are unsuited for studying systems with molecular-scale
inhomogeneities.

@@

\section{Theory and Methods}
We construct a new classical density functional for water.  We
start with a treatment of each individual water molecule as a purely

@@

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional, which
incorporates effects of kinetic energy.  The ideal gas functional is
given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{\frac{n(\xx)}{n_Q}} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules and $n_Q$ is the
quantum concentration
\begin{equation}\label{quantumconcentration}
 n_Q =\left(\frac{mk_BT}{2\pi\hbar^2}\right)^{3/2}.
\end{equation}
The ideal gas free energy functional satisfies the contact density
theorem, which states that the contact density at a hard surface $n_c$
is proportional to the pressure on that surface. This
also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes: 
\begin{equation}\label{contactvaluethm}
  F = n_c k_BT V\:.
\end{equation}
These properties are retained by the total functional,
provided the remaining terms are \emph{purely nonlocal}.  A common
failing of SAFT-based functionals (see, for instance
references\cite{felipe2001examination, gloor2002saft,
  gloor2004accurate, clark2006developing, gloor2007prediction,
  kahl2008modified, gross2009density}) is to treat other terms in the
functional locally, resulting in a functional that is suitable only
for studying the liquid-vapor interface.

@@

%\begin{figure}
%\begin{center}
%\includegraphics[width=\columnwidth]{figs/temperature-versus-density}
%\end{center}
%\caption{The theoretical versus experimental coexistence curve. Away from
%the critical temperature, the agreement is good.\needsworknow{(keep figure?
%its the same as Clark paper)} }
%\label{fig:temperature-vs-density}
%x\end{figure}

\subsection{Liquid-vapor interface}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/surface-298}
\end{center}
\caption{The liquid-vapor density profile at room temperature shows
some small oscillations close to the interface.  }
\label{fig:liquid-vapor-profile}
\end{figure}

We set up a liquid-vapor
interface and plot the density profile.
The fluid density at a liquid-solid interface commonly displays
oscillatory behavior, but the presence of density oscillations at
liquid-vapor interfaces is more debated\cite{penfold2001structure}.
In fact, the \emph{meaning} of the density profile at a liquid-vapor
interface is clouded by the existence of long-wavelength perturbations
known as \emph{capillary waves}.  These waves are not accounted for in
density-functional theory (although the exact
density-functional theory would include them), with the result
that the density profile will be smoothed out by capillary wave effects. 
The liquid-vapor
interface is plotted in Figure~\ref{fig:liquid-vapor-profile} and shows
some small oscillations.

@@

We move into two dimensions by studying a single hydrophobic rod
immersed in water. We use a lattice spacing of about 0.01~nm (0.2 bohrs), 
which is a reasonable resolution based on several lattice spacing calculations.
The results for this spacing of 0.01~nm are within 0.5\% of the free energy 
and a few percent of the contact
density for the highest resolution tested.
Figure~\ref{fig:density-single-rod} shows density
profiles for three different radii (0.3~nm, 0.5~nm, and 1~nm) which have the
expected oscillations outside the rod that reduce to the bulk saturated liquid
density as the distance away from the rods increases. The small radius
rod has comparatively larger oscillations and a higher contact density.
\needsworknow{Expand or add to this description of density profiles and
oscillations for single rod?}

>>>>

% fixme is intended to prioritize stuff that needs fixing.
\newcommand{\fixme}[1]{\textcolor{red}{[\emph{#1}]}}

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, gloor2002saft, Jaqaman2004,
  clark2006developing, lischner2010classical}\fixme{ADD REST OF
  LIST HERE!}, capturing some of the qualitative behavior of water.
However, many of these are less sophisticated than those describing
the hard-sphere fluid, and a quantitatively accurate functional
describing water---which reproduces the temperature dependence of the
macroscopic surface tension, its decrease as curvature increases, and
its quantitative value---has not yet been published.  For instance,
the functional of Lischner \emph{et al}\cite{lischner2010classical}
treats the surface tension correctly, but can only be used at room
temperature.  A numer of classical density functionals have recently
been produced that are based on Statistical Associating Fluid Theory
(SAFT)\cite{gloor2002saft, clark2006developing}\fixme{add
  more}.  These functionals are based on a perturbative thermodynamic
expansion, and give a reasonable prediction for the temperature
dependence of the surface tension.  However, they also overestimate
the surface tension by \fixme{how much? significant factor, I
  think}, and are unsuited for studying systems with molecular-scale
inhomogeneities.

@@

\section{Theory and Methods}
We construct a new classical density functional for water.  We
start with a treatment of each individual water molecule as a purely
classical hard sphere and add association terms based on Statistical 
Associating Fluid Theory (SAFT). Water is unlike simple
fluids (which are mainly defined by van der Waals forces and weak electrostatic
interations) because it requires direct treatment of association effects. The strong 
hydrogen bonds and polar
interactions in water are ideal for describing with SAFT, which was designed for
use with hydrogen-bonding fluids. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules. SAFT can account for the association energies
as well as any effects due to clusters or chains forming \fixme{not true}
in the liquid\cite{muller2001molecular}. SAFT assumes a repulsive hard
sphere core for each atom with a number of short-ranged off center
attractive sites. This model can be applied to give an equation of state for
various systems even inhomogeneous fluids.
\fixme{Expand this a
  bit to accurately describe SAFT and its advantages.}

@@

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional, which
incorporates effects of kinetic energy.  The ideal gas functional is
given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{\frac{n(\xx)}{n_Q}} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules and $n_Q$ is the
quantum concentration
\begin{equation}\label{quantumconcentration}
 n_Q =\left(\frac{mk_BT}{2\pi\hbar^2}\right)^{3/2}.
\end{equation}
The ideal gas free energy functional satisfies the contact density
theorem, which states that the contact density at a hard surface $n_c$
is proportional to the pressure on that surface. This
also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes: 
\begin{equation}\label{contactvaluethm}
  F = n k_BT V\:.
\end{equation}\fixme{may be ambiguous about n}
These properties are retained by the total functional,
provided the remaining terms are \emph{purely nonlocal}.  A common
failing of SAFT-based functionals (see, for instance
references\cite{felipe2001examination, gloor2002saft,
  gloor2004accurate, clark2006developing, gloor2007prediction,
  kahl2008modified, gross2009density}) is to treat other terms in the
functional locally, resulting in a functional that is suitable only
for studying the liquid-vapor interface.

@@

%\begin{figure}
%\begin{center}
%\includegraphics[width=\columnwidth]{figs/temperature-versus-density}
%\end{center}
%\caption{The theoretical versus experimental coexistence curve. Away from
%the critical temperature, the agreement is good.\fixme{(keep figure?
%its the same as Clark paper)} }
%\label{fig:temperature-vs-density}
%x\end{figure}

The fluid density at a liquid-solid interface commonly displays
oscillatory behavior, but the presence of density oscillations at
liquid-vapor interfaces is more debated\cite{penfold2001structure}.
In fact, the \emph{meaning} of the density profile at a liquid-vapor
interface is clouded by the existence of long-wavelength perturbations
known as \emph{capillary waves}.  These waves are not accounted for in
density-functional theory (although the exact
density-functional theory would include them), with the result
that the density profile will be smoothed out by capillary wave
effects. We set up a liquid-vapor interface and observed small density
oscillations in the water near the surface.

@@

We move into two dimensions by studying a single hydrophobic rod
immersed in water. We use a lattice spacing of about 0.01~nm (0.2 bohrs), 
which is a reasonable resolution based on several lattice spacing calculations.
The results for this spacing of 0.01~nm are within 0.5\% of the free energy 
and a few percent of the contact
density for the highest resolution tested.
Figure~\ref{fig:density-single-rod} shows density
profiles for three different radii (0.3~nm, 0.5~nm, and 1~nm) which have the
expected oscillations outside the rod that reduce to the bulk saturated liquid
density as the distance away from the rods increases. The small radius
rod has comparatively larger oscillations and a higher contact density.
\fixme{Expand or add to this description of density profiles and
oscillations for single rod?}
